---
title: "Pick a team"
author: "Tom Conway"
format: 
  gfm: default
  html:
    embed-resources: true
editor: visual
---

## Pick a team

-   Import player data and set up packages

```{r}

#|label: import-points-predictions

library(tidyverse)
library(tidymodels)

player_data <- read_csv("processed_data/player_data.csv")

```

-   Create and model algorithm for picking players to make up a 15
-   Create a function to assess if a team is legitimate or not, its cost, and its likely score

```{r}

#| label: team-functions

example_team <- player_data |> filter(
  web_name %in% c(
    "Pickford",
    "Sels",
    "Gvardiol",
    "Murillo",
    "Mings",
    "Pinnock",
    "Aina",
    "M.Salah",
    "Joelinton",
    "Rogers",
    "J.Murphy",
    "Kluivert",
    "Bowen",
    "Wissa",
    "Mateta"
  )
)
  
# Check if it's a legal team (not cost)

team_checker <- function(team) {
  is_15 <- nrow(team) == 15
  is_2_GK <- nrow(team |> filter (element_type == 1)) == 2
  is_5_DF <- nrow(team |> filter (element_type == 2)) == 5
  is_5_MD <- nrow(team |> filter (element_type == 3)) == 5
  is_3_FW <- nrow(team |> filter (element_type == 4)) == 3
  teams_with_over_3 <- count(team, short_name, sort = TRUE) |>
    filter(n > 3)
  max_three_per_team <- nrow(teams_with_over_3) == 0
  all(is_15, is_2_GK, is_5_DF, is_5_MD, is_3_FW, max_three_per_team)
}

team_checker(example_team)

# Check cost

team_cost <- function(team) {
  team |> summarise(total = sum(now_cost, na.rm = TRUE)) |> pull(total)

}

team_cost(example_team)

# Calculate team score

team_score <- function(team) {
  # Set out likely weights for play time (can optimise later)
  weights <- tribble(
    ~weight, 0.7, 0.3, 1, 1, 0.7, 0.7, 0.6, 1.5, 1.3, 1, 0.6, 0.4, 1.2, 0.6, 0.4
  )
  #organise to be in the right order for weighting
  team|>
    arrange(desc(predicted_2026_points)) |>
    arrange(element_type) |>
  # then weight according to likely play time and captaincy and sum
    cbind(weights)|>
    mutate(weighted_score = predicted_2026_points * weights) |>
    summarise(total = sum(weighted_score))|>
    pull()
}

team_score(example_team)
```

-   Eliminate dominated players
-   Use a greedy algorithm with cost penalty and randomness to create seeds (first of all need a checker of is it not broken yet - dummy players with minimum costs, fake teams, and 0 score)
-   Do some swaps to see if you can improve on the seeds (how best to?)
